---
# output: bookdown::pdf_document2
output: bookdown::github_document2
title: "Distance decay functions for modelling active travel uptake"
bibliography: 
  - ddp.bib
  - /home/robin/uaf/allrefs.bib
  # - ddp2.bib]
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, eval=FALSE, echo=FALSE}
citr::tidy_bib_file(rmd_file = "README.Rmd", messy_bibliography = "~/uaf/allrefs.bib", file = "ddp2.bib")
browseURL("README.pdf")
```


```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
library(tidyverse)
by_city_lm = readRDS("by_city_lm.Rds")
by_city_beta = readRDS("by_city_beta.Rds")
```

<!-- paper highlights -->
<!-- many ways to simulate the impact of distance -->
<!-- pros and cons of different approaches -->
<!-- for walking:  -->
<!-- for cycling: -->
<!-- next steps... -->

# Abstract

# Introduction

Trips of short distances tend to be more frequent than trips of long
distances. This is a fundamental feature of transport behaviour,
observed in the majority of transport systems worldwide. The concept is
neatly captured by the term *distance decay* (*dd*), meaning simply that
the proportion of trips of any given type tends to zero as distance
tends to infinity.
Distance is an especially important variable for walking and cycling trips due to their low speeds compared with many motorised modes (outside rush hour congestion at least), meaning that that the concept of distance decay is especially important for active modes.
However, notwithstanding a few notable exceptions, the concept has not been explored in depth from walking and cycling perspectives [@Iacono2008;@millward_activetransport_2013;@rybarczyk_spatial_2018].

This paper aims to fill a gap in the literature by providing an overview of distance decay functions that have been used in the literature --- and some that have not --- from the perspective of potential applications to active travel research.
It is structured as follows.
...
Section \@ref(funs) contains the bulk of the analysis, demonstrating a methods for implementing a range of distance decay functions.
A summary of results is presented in Section \@ref(results), which are discussed in Section \@ref(discussion-and-conclusion).

# Defining distance decay

*Distance decay* --- along with the synonymous *deterrence function*
[@Simini2012] --- builds on older terms expressing the same fundamental
truth, such as the 'friction of distance' and the 'gravity model'
(described in the subsequent section). Work focussed explicitly on $dd$
explores *how* the proportion of trips ($p$) declines with distance
($d$) [@Iacono2008], beyond the simple observation that the slope is
negative beyond a certain threshold distance. A resergence of interest
in the distance-frequency relationship has refined and, in some cases,
sought to overturn pre-existing (and often context-specific)
formulations, such as $p = \alpha e^{\beta d}$ [@Wilson1971] and $p =
\alpha d^{\beta}$ [@Fotheringham1981]. Much *dd* research asks: what
is the best function for linking the proportion (different types of)
trip to distance?

Mathematically, what is *f* of *x*

$$ p = f(x) $$

```{r, echo=FALSE}
# Look into changing this to a partial variable 
# Convert ds in xs
# $$
# \frac{\delta p}{\delta d} < 0 
# $$
# where $d$ is distance (km), $X$ is a set of explanatory variables, $tr$
# is the trip rate, expressed either in absolute numbers or as a proportion of
# all trips for a given trip distance and $e$ is stochastic error.
```

where $x$ is distance and $p$ is the proportion of trips made in any
distance band? Distance decay models can be empirically tested, and
therefore ranked, by comparing observed travel behaviour with the levels
expected based on different distance decay functions.

Equation 1.1 can clearly be refined in many directions. Subscripts can
be added to the terms and new variables can be added to illustrate how
distance decay changes with a range of additional factors such as type
of bicycle and hillness of the route. how distance decay changes in
response to a range of variables. The type of trip (e.g. time of day,
purpose), characteristics of the people making the trip (e.g. age and
sex) and the location and physical surroundings of the trip (e.g.
hilliness, transport infrastructure) have all been found to affect the
shape of distance decay curves
[@Fotheringham1981;@Fingleton2005].
Critically for understanding transport systems is the *mode*, or
'method' of transport. Mode refers to the choice of 'vehicle' (e.g.
walking, cycling, car driving) and has been found, unsurprisingly, to
dramatically affect the speed travelled during and amount of effort
required for trips of different distances. The *dd* concept is
especially applicable to active travel modes due to physiological limits
on human mechanical power output and the slow speed of these modes. The
next section shows that the concept has a long history in the academic
literature but that interest in the term from the perspective of walking
and cycling is relatively recent.

```{r, echo=FALSE}
# Todo: add some theoretical dd curves here
```

# The distance decay literature

A variety of terms have been used to express the idea underlying the
$dd$ concept outlined in this paper. These including the 'first law of
geography', the 'friction of distance' and 'the gravity model'. It is
worth reviewing these terms before describing work refering to $dd$
directly, to put the term in its wider historical context. The final
part of this section reviews recent literature explicitly using the term
'distance decay' to explore the concept

## The gravity model

The 'gravity model' of movement patterns helped to quantify and
generalise early incarnations if *dd* [@zipf1946p1]. This rule (it is
sometimes referred to as the 'gravity law') suggests that travel
incentives are roughly analogous to Newtonian gravitation
[@Ravenstein1885]. The resulting formulae imply that the total movement
between two places ($T_{ij}$ between origin $i$ and destination $j$) is
proportional to the product of their populations ($m_i$ and $n_j$),
divided by a power of the distance between them:

$$
T_{ij} = \frac{m_i n_j}{d^\beta}
$$

as it has sometimes been called has been a rich source of theoretical
and methodological advance in many fields, primarily urban modelling but
also in fields as diverse as highway planning [@Jung2008], national
transport of minerals [@Zuo2013] and spatial epidemiology
[@Balcan2009].

## The radiation model

Despite dissenting voices --- including the statement that "a strict
gravity model simply did not work" for modelling urban systems and that
some subsequent refinements to the gravity model were "fudge factors"
[@Wilson1998-past] --- the gravity model has been one of the dominant
tools for understanding urban mobility over the past 100 years
[@masucci2013gravity]. A recent development in this field has been the
'radiation model', which has been found to fit travel to work and other
flow data well [@Simini2012]. This new formula for estimating flow
rates between geographic zones is interesting in its omission of
distance as an explicit explanatory variable. Instead, the radiation
model uses the number 'intervening opportunities' ($IO$) as a proxy for
$dd$ the denominator to estimate flow:

$$ dd \approx \frac{m_in_i}{(m_i+s_{ij})(m_i+n_j+s_{ij})} $$

where $m$ and $n$ are origin and destination zones whose population (or other measure of 'size') is iterated through for each origin $i$ and each destination $j$, and where $s_{ij}$ represents the total population (excluding the origin and destination zones) within a radius $r_{ij}$ equal to the distance between zones $i$ and $j$. 

An advantage of the approach is its 'parameter free' nature; a disadvantage is the additional computation needed to calculate $s_{ij}$.
Unlike other distance decay approaches, the distance alone is insufficient to estimate the amount of travel.
A recent study compared the parameter-free radiation model against the
gravity model on a large intra-city (London) dataset on commuting. It
neither model produced a satisfactory fit with the data,
leading to the conclusion that "commuting at the city scale still lacks
a valid model and that further research is required to understand the
mechanism behind urban mobility" [@masucci2013gravity].
The 'first law
of geography' and the related concept of the 'friction of distance' are
alternative yet closely related (i.e. not mutually exclusive) terms for
exploring $dd$.

## The first law of geography

Tobler's famous **first law of geography** states that "everything is
related to everything else, but near things are more related than
distant things" [@Tobler1970]. The phrase implies that interaction
between things declines with distance without specifying how or why.
Clearly, the increased frequency of communications and transport between
places that are close to each other can help explain spatial
autocorrelation [@Miller2004]. However, unlike $dd$, the 'first law'
says little about the way in which relatedness between entities declines
with distace. Moreover, in a world of accelerating globalisation under
the auspices of the ongoing 'digital revolution', the relevance of the
'first law' to the system-level processes it was proposed to explain has
come under scrutiny [@Westlund2013].

```{r, echo=FALSE}
# and related quantitative measures of Tobler's law
```

To overcome this issue, the scope of this paper is limited to
transportation of people and goods, opposed to immaterial communication.
This simplifies the issue and makes it more tangible. Due to fundament
physical limits on the efficiency with which matter can be moved
[@MacKay13082013], and a limited supply of energy (especially pertinent
in an era of resource over-extraction and climate change), such physical
transport will always be limited to some degree by distance. This notion
of travel frequency tending to zero as distance tends to infinity,
present in the 'friction of distance' terminology, is also encapsulated
by the more recent phrase 'distance decay' used in this paper.

## The friction of distance

The concept of the 'friction of distance' has been in use for over 100
enjoying steady (albeit slowing) growth in use until the early 2000s
(Fig. 1). The term helps to explain why Tobler's Law is true, implying
that it is more *difficult* (e.g. in terms of energy use) for distant
things to interact. 'Friction' is thus defined as "the difficulty of
moving a volume (passengers or goods)" [@Muvingi2012]. The friction of
distance is predominantly used as a synonym for $dd$, and to some extent
the slowing use of the term illustrated in Fig. 1 can be seen as a
consequence of the latter simply replacing the former. One study, for
example, refers to "the friction of distance parameter in gravity
models" [@cliff1974evaluating] whereas others refer to 'distance decay'
in spatial interaction and gravity models respectively: the terms are
largely interchangeable [@Griffith1980]. However, the
question of precisely *how* the metaphorical friction increases with
distance has been tackled to a lesser extent in the (generally older)
literature using the term. The recent $dd$ literature, by contrast, is
largely focussed this question, as described in the subsequent section.

## Recent distance decay literature

As illustrated in Figure \ref{fcitations}, $dd$ has grown rapidly as a
term in the academic literature over the past 50 years, compared with
the well-established "gravity model" terminology. By the 1970s, it seems
that $dd$ overtook the "friction of distance" amongst transport
researchers. Although Tobler's 'First Law' of geography has gained rapid
acceptance since its inception in the 1970s, it has a far lower (by a
factor of 5) rate of use than $dd$ in the transport literature. In other
words, a quantitative review of the literature demonstrates that $dd$ is
an up-and-coming term in transport studies. This section does not
attempt a full overview of the 10,000+ articles using the $dd$
terminology in the academic literature, instead focussing on a few key
and highly cited works that helped set the agenda for $dd$ research.

"friction of distance" terms. (which tends to assign importance to
interaction between places rather than the impact of distance)

```{r, echo=FALSE, fig.cap="Number of citations per decade for different terms for 'distance decay' from Google Scholar.", fig.lp="fcitations"}
brt <-  theme(axis.text.x = element_text(angle=20, color = "black"),
        panel.background = element_rect(fill = "white"),
        panel.grid.major.y = element_line(color = "grey", linetype=3))
ddt <- read.csv("https://github.com/npct/pct-team/raw/master/input-data/dd-terms-scholar.csv")
ddt$Term <- gsub(pattern = "First law of geography", replacement = "First law", ddt$Term)
ggplot(ddt) +
  geom_line(aes(x = Decade, y = Frequency, color = Term, group = Term)) +
  scale_y_log10() + brt +
  ylab("Citations in Google Scholar")
# +
#   theme(legend.position = "top")
# ggsave("paper/fcitations.pdf", scale = 1.5)
```

Within the general understanding that trips become less frequent with
distance (beyond some threshold limit) there are various ways of
describing the dependent variable that *dd* functions seek to explain.
*dd* can be understood as:

-   the absolute number of trips expected for any given distance band
    ($db$): $f(d) = T_{db}$
-   the proportion of *all* trips that are made for a given distance
    band: $f(d) = T_{db} / T$
-   the proportion of trips *within a given distance band* made by a
    particular trip type (e.g. walking): \newline $f(d) = Twalk_{db} /
    T_{db}$

Of these definitions the second is the most generalisable, being a
proportion ($0 \geq p \leq 1$) that tends to zero with increasing
distance for any mode of transport. The third definition of *dd* also
estimates a proportion with the same constraints, but is less
generalisable: the result will not always tend to zero (the proportion
of trips by air travel, for example, will tend to one for large trip
distances). The first and simplest definition is the least generalisable
as it is highly dependent on the total amount of travel.

# Distance decay functions {#funs}

Four functional forms commonly used in the literature to characterise
distance decay curves were described in @martinez_new_2013 and @levinson_transport_2020.
<!-- [@Martinez2013]  -->
as:

-   Exponential functions, $e^{\beta x}$ [@Wilson1971].
-   Power functions, $x^{\beta}$ [@Fotheringham1981].
-   Tanner functions, $x^{\beta_1} e^{\beta_2 x}$ [@Openshaw1977].
-   Box-Cox functions, $exp(\beta \frac{x^{\gamma} - 1}{\gamma})$
    when the parameter $\gamma \neq 0$ and $x^{\beta}$ when $\gamma
    = 0$

In addition to these, we have added the additional functions:

- Modified beta distributions, which have been used to model distance decay in ecological research [@nekola_scale_2014]
- A modified version of the generalised logistic or 'Richards' function

$$ f(x) = \frac{1}{(1 + Qe^{-B(x - M)})^{1/v}} $$

Each of the 4 parameters in this model can be altered to ensure optimal
fit and each has a separate meaning:

-   $B$ is analogous to $\beta$ in the other functions
-   $Q$ represents the intercept
-   $v$ affects the skewness of the curve
-   $M$ is the *x* value of maximum growth

Each of these distance decay functions was implemented on data representing commuting patterns by walking and cycling in three major UK cities with moderate, low and very low levels of walking and cycling mode share compared with international best practice: Cambridge, Leeds, and the rural local authority of Hereford.
These cities were selected because they are of similar size but have widely different levels of walking and cycling, as shown in Figure \@ref(fig:pwalkcycle).
The aim is to test the distance decay functions in diverse contexts.
Data on travel behaviour and route distances was taken from the Propensity to Cycle Tool, which provides open data on cycling and other modes across England from the 2011 Census [@lovelace_propensity_2017].

```{r, echo=FALSE, eval=FALSE}

```

```{r pwalkcycle, fig.cap="The proportion of people who walk (above) and cycle (below) to work in major (50+ commuters) origin-destination pairs in three English regions: Cambridge (left), Avon (centre) and West Yorkshire (right)."}

```

## Linear models

The simplest model to fit to the data is a linear model. For
illustrative purposes we will fit a linear model to the data presented
in Figure 1, with different intercepts and gradients for each of the
groups (Figure 2).


The intercepts and gradients for each group are presented in table !!!

The above numbers are equations that describe the relationship between
distance and *clc* for each group. In the `Mal_Young_NC` group, for
example.

```{r linear, fig.cap="Simple linear model."}
ggplot(by_city_lm) +
  geom_point(aes(distance, proportion, size = all), alpha = 0.1) +
  facet_grid(mode ~ region) +
  geom_line(aes(distance, pred), col = "red", size = 2) +
  ylim(c(0, 1)) +
  xlim(c(0, 10))
```


## Exponential functions


## The beta function

```{r beta, fig.cap="The beta function."}
ggplot(by_city_beta) +
  geom_point(aes(distance, proportion, size = all), alpha = 0.1) +
  facet_grid(mode ~ region) +
  geom_line(aes(distance, pred), col = "red", size = 2) +
  ylim(c(0, 1)) +
  xlim(c(0, 10))
```




# Results


# Discussion and conclusion


# References
